Did WW2 Produce an Equivalent to a Modern Computer?

### 2017-06-12 Draft 2/Section 2 - Turing Machines.tex
*Tick Tock Tick Tock* 6 days to goooo yayyy
Yeh I did loads of changes and things, look at the log if you REALLY care that much

\lstdefinelanguage{Turlang}
{
  morekeywords={
    function,
    bool,
    int,
    and,
    or,
    not,
    xor
  },
  sensitive=true, % keywords are not case-sensitive
  morecomment=[l]{//}, % l is for line comment
  morecomment=[s]{/*}{*/}, % s is for start and end delimiter
  morestring=[b]" % defines that strings are enclosed in double quotes
}
\lstset{language=Turlang}
\begin{document}
\renewcommand{\arraystretch}{1.5}

At roughly the same time, the English mathematician Alan Turing was constructing a method to achieve the same goal, but in an entirely different way. Rather than building a basic and strongly theoretical system, the concept used was far more physical and far closer to the mechanical computers that eventually implemented this idea.

\subsection{Examples of simple Turing Machines}\label{sec:TMacIntro}
The Turing Machine is composed of two parts. Firstly, an ordered set of \textit{cells} of infinite\footnote{Many definitions of Turing Machines have the tape extending infinitely in one direction only - in fact for full functionality a tape extending infinitely in both directions is unnecessary.} length known as \textit{tape}. On each of these cells one of a finite set of \textit{symbols} - either alphanumeric characters or the symbol which represents an empty cell. At a given point on this tape is a \textit{tape head} (or cursor) which is able to write symbols to the current cell. Secondly, a finite set of instructions known as \textit{states} contain finite-length sequences of operations known as the \textit{transition function} which the cursor enacts, and the states that the machine moves to after performing those operations. An example description of a set of states would be: \cite{turing1936computablenumbers}

@@

This table format shall be used to describe Turing Machine configurations. First of all, the last line describes the starting state ($\mathbb{A}$). When the machine begins an iteration on any one of these states, it first checks if the symbol aspect of the configuration is equal to the \textit{symbol} currently under the cursor on the tape. If this condition is met, the transition function is carried out - P\textit{n} for any symbol \textit{n} prints the symbol \textit{n} on the tape under the cursor; E erases any symbol under the cursor\footnote{More formally, prints the empty cell symbol}; L moves the cursor to the left; R moves the cursor to the right. When these operations are completed, the state is set to that described in the last column. The machine then moves to the next iteration.

@@

The operation of such a machine is significantly more complex than the earlier one. Key to the operation is the fact that cells are paired into value cells (odd-numbered cells) on which the symbols of the output sequence are printed. Only the symbols \texttt{0} and \texttt{1} will ever be printed on these cells. To the right of each value cell is a label cell (an even-numbered cell). These cells will either be left empty or will contain a 'marker' symbol, which labels the symbol on the value cell to the left. These symbols can be erased and overwritten, and are not part of the output sequence. Only one label cell is needed to contain multiple markers, or more descriptive markers for a single value cell, as this can be achieved with a sufficiently rich variety of marker symbols. This method is frequently used by Alan Turing in \cite{turing1936computablenumbers}.

This machine operates by first printing the symbols \texttt{e e 0} at the start of the tape, before printing the first \texttt{0} of the sequence on the fifth cell, finishing on the third cell and moving to state $\mathbb{A}$. (in future I shall represent this finishing state as $\rightarrow \mathbb{A}$).

If $\mathbb{A}$ is triggered, every \texttt{1} consecutively to the left of the cursor inclusively will be marked with \texttt{x}, $\rightarrow \mathbb{B}$.

If $\mathbb{B}$ is triggered, the symbol \texttt{1} is appended to the output sequence, $\rightarrow \mathbb{C}$.

If $\mathbb{C}$ is triggered, the cursor moves leftwards through the labels: if an \texttt{x} is found it is erased and $\rightarrow \mathbb{B}$; if the cursor reaches the beginning (an \texttt{e} is found) $\rightarrow \mathbb{D}$.

If $\mathbb{D}$ is triggered, the symbol \texttt{0} is appended to the output sequence, the cursor moves to the previous value cell and $\rightarrow \mathbb{A}$.

The workings of this algorithm are mostly in states $\mathbb{A}$ and $\mathbb{C}$: after the \texttt{0} is drawn $\mathbb{A}$ draws a number of \texttt{x} symbols equal to the previous number of \texttt{1} symbols drawn. $\mathbb{C}$ triggers $\mathbb{B}$ a number of times equal to the number of found \texttt{x} symbols before drawing a \texttt{0} - as $\mathbb{B}$ has already been triggered once, a number of \texttt{1} symbols will be drawn which is one more than the previous number drawn. For clarity in the comprehension of this somewhat abstract algorithm, some of the early states have been listed on the next page. Empty squares are represented by dots ($\cdot$) and the tape states continue in columns. An ellipsis marks the omission of a set of states which are implied or trivial.

\begin{equation*}
\begin{aligned}
&\mathtt{e\ e\ \underset{\uparrow}{0}\ \cdot\ 0} & \rightarrow \mathbb{A}\ \ \ \ \ \ \ \ \ \ & ...&\\
&\mathtt{e\ e\ \underset{\uparrow}{0}\ \cdot\ 0} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ \underset{\uparrow}{\cdot}} & \rightarrow \mathbb{D}\\
&\mathtt{e\ e\ 0\ \cdot\ \underset{\uparrow}{0}} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ \underset{\uparrow}{1}\ \cdot\ 0} & \rightarrow \mathbb{A}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ \underset{\uparrow}{\cdot}} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ \underset{\uparrow}{1}\ \cdot\ 1\ x\ 0} & \rightarrow \mathbb{A}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \underset{\uparrow}{\cdot}\ 1} & \rightarrow \mathbb{C}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ \underset{\uparrow}{0}\ \cdot\ 1\ x\ 1\ x\ 0} & \rightarrow \mathbb{A}\\
&\mathtt{e\ e\ 0\ \underset{\uparrow}{\cdot}\ 0\ \cdot\ 1} & \rightarrow \mathbb{C}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ \underset{\uparrow}{0}\ \cdot\ 1\ x\ 1\ x\ 0} & \rightarrow \mathbb{B}\\
&\mathtt{e\ \underset{\uparrow}{e}\ 0\ \cdot\ 0\ \cdot\ 1} & \rightarrow \mathbb{C}\ \ \ \ \ \ \ \ \ \ & ...&\\
&\mathtt{e\ e\ \underset{\uparrow}{0}\ \cdot\ 0\ \cdot\ 1} & \rightarrow \mathbb{D}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ x\ 1\ x\ 0\ \cdot\ \underset{\uparrow}{\cdot}} & \rightarrow \mathbb{B}\\
&\mathtt{e\ e\ 0\ \cdot\ \underset{\uparrow}{0}\ \cdot\ 1} & \rightarrow \mathbb{D}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ x\ 1\ x\ 0\ \underset{\uparrow}{\cdot}\ 1} & \rightarrow \mathbb{C}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ \underset{\uparrow}{1}} & \rightarrow \mathbb{D}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ x\ 1\ \underset{\uparrow}{x}\ 0\ \cdot\ 1} & \rightarrow \mathbb{C}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ \underset{\uparrow}{\cdot}} & \rightarrow \mathbb{D}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ x\ 1\ \cdot\ \underset{\uparrow}{0}\ \cdot\ 1} & \rightarrow \mathbb{B}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ \underset{\uparrow}{1}\ \cdot\ 0} & \rightarrow \mathbb{A}\ \ \ \ \ \ \ \ \ \ & ...&\\
&\mathtt{e\ e\ 0\ \cdot\ \underset{\uparrow}{0}\ \cdot\ 1\ x\ 0} & \rightarrow \mathbb{A}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ x\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ \underset{\uparrow}{\cdot}} & \rightarrow \mathbb{B}\\
&\mathtt{e\ e\ 0\ \cdot\ \underset{\uparrow}{0}\ \cdot\ 1\ x\ 0} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ x\ 1\ \cdot\ 0\ \cdot\ 1\ \underset{\uparrow}{\cdot}\ 1} & \rightarrow \mathbb{C}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ \underset{\uparrow}{1}\ x\ 0} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & ...&\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ x\ \underset{\uparrow}{0}} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \underset{\uparrow}{x}\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1} & \rightarrow \mathbb{C}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ x\ 0\ \cdot\ \underset{\uparrow}{\cdot}} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ \underset{\uparrow}{1}\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1} & \rightarrow \mathbb{B}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ x\ 0\ \underset{\uparrow}{\cdot}\ 1} & \rightarrow \mathbb{C}\ \ \ \ \ \ \ \ \ \ & ...&\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \underset{\uparrow}{x}\ 0\ \cdot\ 1} & \rightarrow \mathbb{C}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ \underset{\uparrow}{\cdot}} & \rightarrow \mathbb{B}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ \underset{\uparrow}{0}\ \cdot\ 1} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \underset{\uparrow}{\cdot}\ 1} & \rightarrow \mathbb{C}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ \underset{\uparrow}{1}} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & ...&\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ \underset{\uparrow}{\cdot}} & \rightarrow \mathbb{B}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ \underset{\uparrow}{e}\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ 1} & \rightarrow \mathbb{C}\\
&\mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \underset{\uparrow}{\cdot}\ 1} & \rightarrow \mathbb{C}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ \underset{\uparrow}{0}\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ 1} & \rightarrow \mathbb{D}\\
&...& &...&\\
&\mathtt{e\ \underset{\uparrow}{e}\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1} & \rightarrow \mathbb{C}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ 1\ \cdot\ \underset{\uparrow}{\cdot}} & \rightarrow \mathbb{D}\\
&\mathtt{e\ e\ \underset{\uparrow}{0}\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1} & \rightarrow \mathbb{D}\ \ \ \ \ \ \ \ \ \ & \mathtt{e\ e\ 0\ \cdot\ 0\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ 0\ \cdot\ 1\ \cdot\ 1\ \cdot\ \underset{\uparrow}{1}\ \cdot\ 0} & \rightarrow \mathbb{D}\\
\end{aligned}
\end{equation*}

\clearpage

It is clear from this example that the operation of the Turing Machine is very different from that of an ordinary computer. However, the core components are still mostly present - conditionals in the form of the symbol tests in the second column, variables in the form of labels, and continuity and looping when a state triggers itself - a form of recursion. The end result is the same - a solution is produced in the form of the computable sequence written on the value squares, and the machine can continue this sequence without halting indefinitely. Unfortunately, the operation of this machine is somewhat complex and involved. To simplify this, we will introduce a type of function for the Turing machine.

@@

\textit{M}-functions are effectively an extension of states described by Turing in \cite{turing1936computablenumbers}. They are denoted by a lowercase letter and a set of arguments contained in brackets - arguments in gothic script refer to states (or other \textit{m}-functions), and arguments in italics refer to symbols. When an \textit{m}-function is triggered, the italic and gothic symbols in the description are substituted by the supplied arguments (similar to application in $\lambda$-Calculus). An example of an \textit{m}-function would be the function $\mathbb{f}$ below: \cite{turing1936computablenumbers}

@@

The function $\mathbb{f}$ takes three arguments - the states \textfrak{C} and \textfrak{E} and the symbol \textit{a}. It is not difficult to see that if $\rightarrow \mathbb{f}$($\mathbb{A}$,$\mathbb{B}$,x) the cursor would trigger $\mathbb{A}$ at the leftmost occurrence of \texttt{x}. If no symbols \texttt{x} were found, \textfrak{E} would be triggered on the marker of the first empty value square.

@@

\medskip\noindent\begin{tabu} to \textwidth{XXXXX}
    \multicolumn{2}{c}{\textit{Configuration}} & \multicolumn{2}{c}{\textit{Behavior}} \\
    \textit{state} & \textit{symbol} & \textit{transition function} & \textit{final state} \\
    \hhline{====}
    $\mathbb{e}$(\textit{a}, \textfrak{C}, \textfrak{E}) & - & - & $\mathbb{f}$(\textit{a}, $\mathbb{e}_1$(\textfrak{C}), \textfrak{E}) & \multirow{2}{80pt}{Erases first occurrence of symbol \textit{a}} \\
    \hhline{----}
    $\mathbb{e}_1$(\textfrak{C}, \textfrak{E}) & - & E & \textfrak{C}\\
    \hhline{====}
    $\mathbb{l}$(\textfrak{C}, \textfrak{E}) & - & L & \textfrak{C} & Moves cursor left\\
    \hhline{----}
    $\mathbb{r}$(\textfrak{C}, \textfrak{E}) & - & R & \textfrak{C} & Moves cursor right\\
    \hhline{====}
    $\mathbb{fl}$(\textit{a}, \textfrak{C}, \textfrak{E}) & - & - & $\mathbb{f}$(\textit{a}, $\mathbb{l}$(\textfrak{C}), \textfrak{E}) & Moves to left of \textit{a} \\
    \hhline{----}
    $\mathbb{fr}$(\textit{a}, \textfrak{C}, \textfrak{E}) & - & - & $\mathbb{f}$(\textit{a}, $\mathbb{r}$(\textfrak{C}), \textfrak{E}) & Moves to right of \textit{a} \\
    \hhline{====}
    $\mathbb{p0}$(\textfrak{C}, \textfrak{E}) & - & P\texttt{0} & \textfrak{C} & Prints \texttt{0} \\
    \hhline{----}
    $\mathbb{p1}$(\textfrak{C}, \textfrak{E}) & - & P\texttt{1} & \textfrak{C} & Prints \texttt{1} \\
    \hhline{----}
    $\mathbb{ps}$(\textit{s},\textfrak{C}, \textfrak{E}) & - & P\textit{s} & \textfrak{C} & Prints \textit{s} \\
    \hhline{====}
    $\mathbb{n}$(\textfrak{C}, \textfrak{E}) & - & - & $\mathbb{f}$(\texttt{e}, $\mathbb{n}_1$(\textfrak{C}), \textfrak{E}) & \multirow{3}{80pt}{Moves to first empty (final) \textit{p}-square} \\
    \hhline{----}
    \multirow{2}{*}{$\mathbb{n}_1$(\textfrak{C})} & Any  & R, R & $\mathbb{n}_1$(\textfrak{C}) \\
                                                  & None & - & \textfrak{C} \\
    \hhline{====}
\end{tabu}

@@

At this point, the similarity between Turing machines and modern computers becomes very clear. The machine is able to store more complex information, from integers to strings of characters to images, as binary sequences of value square \texttt{0}s and \texttt{1}s - these shall be stored on the tape as memory. The first data structure we shall define is the simple boolean - only 1 bit (1 value square) long. In order to identify these sequences on the tape as booleans, we shall label the first character of the boolean with an identifier (we shall call these pointers) unique to that boolean variable - for example the symbol \texttt{\textbf{B}myvar} (bear in mind this is one symbol, not six) would mark a boolean named \textit{myvar}. Let us define some \textit{m}-functions to define and assign values to variables of the boolean type:

@@

\subsection{Scripted Configurations}
As we have moved into more and more complex Turing machine functions, it can be noted that the middle two columns of the description are becoming gradually more scarce - this is due to the fact that their operation can be entirely represented within \textit{m}-functions. We have started doing so with the functions $\mathbb{l}$, $\mathbb{r}$, $\mathbb{p0}$, and $\mathbb{p1}$ - furthermore, it is not difficult to create conditional functions to replicate the operation of the second column.

@@

The function $\mathbb{if}$ takes a symbol argument, which is compared to the symbol under the cursor - if they are the same, the machine $\rightarrow$ \textfrak{O}; if not the machine $\rightarrow$ \textfrak{O}.
Since we have now entirely defined the operation of the machine in terms of \textit{m}-functions, we no longer have any need for the central two table columns. In continuing to define \textit{m}-functions, we could simply choose to omit these columns, or even omit the table altogether and use a clearer notation - in this case, for a definition of the if function but based on pointers:

@@

\subsection{Conclusion}

Turing machines are truly remarkable entities. While $\lambda$-Calculus was able to mimic the operation of a modern computer (behaving like a purely functional programming language such as Haskell), its beginnings were completely abstract - the mathematical operations of abstraction and application, while being probably more useful than Turing machines in yielding results, were unlike anything which could actually be created in the real world. However, Turing machines are completely grounded in physical reality, and it is totally conceivable how one could create such a machine - despite this, they are able to produce results very similar to that of an imperative programming language such as Java - data storage, typed and referenceable variables, and boolean logic, all performed as a sequence of commands. From these building blocks, integers and more complex concepts can be produced. This is the truly remarkable thing about Turing machines - they illustrate clearly how mechanisms can in fact solve computable problems. As such, it is most useful to define a computer as something which is able to behave like a Turing machine - in effect, a computer is distinguished by the fact that it can simulate a Turing machine. A machine with this ability shall be called 'Turing-complete'

>>>>

\lstdefinelanguage{Turlang}
{
  morekeywords={
    function,
    bool,
    int,
    and,
    or,
    not,
    xor
  },
  sensitive=true, % keywords are not case-sensitive
  morecomment=[l]{//}, % l is for line comment
  morecomment=[s]{/*}{*/}, % s is for start and end delimiter
  morestring=[b]" % defines that strings are enclosed in double quotes
}
\lstset{language=Turlang}

\begin{document}

At roughly the same time, the English mathematician Alan Turing was constructing a totally different method of achieving the same goal: rather than a basic and heavily theoretical system, it was far more physical and far closer to the mechanical computers that eventually implemented the idea.

\subsection{Examples of simple Turing Machines}\label{sec:TMacIntro}
The Turing Machine is composed of two parts. Firstly, an ordered set of \textit{cells} of infinite\footnote{Many definitions of Turing Machines have the tape extending infinitely in one direction only - in fact for full functionality a tape extending infinitely in both directions is unnecessary.} length known as \textit{tape}. On each of these cells is one of a finite set of \textit{symbols} - either alphanumeric characters or the symbol which represents an empty cell. At a given point on this tape is a \textit{tape head} (or cursor) which is able to write symbols to the current cell. Secondly, a finite set of instructions known as \textit{states} contain finite-length sequences of operations known as the \textit{transition function} which the cursor enacts, and the states that the machine moves to after performing those operations. An example \textit{description} of a set of states would be: \cite{turing1936computablenumbers}

@@

This description table format shall be used to describe Turing Machine configurations. First of all, the final line describes the starting state ($\mathbb{A}$). When the machine begins an iteration on any one of these states, it first checks if the symbol aspect of the configuration is equal to the symbol currently under the cursor on the tape. If this condition is met, the transition function is carried out - P\textit{n} for any symbol \textit{n} prints the symbol \textit{n} on the tape under the cursor; E erases any symbol under the cursor\footnote{More formally, prints the empty cell symbol}; L moves the cursor to the left; R moves the cursor to the right. When these operations are completed, the state is set to that described in the last column (in future I shall describe this using the $\rightarrow$ symbol, for example $\rightarrow \mathbb{B}$). The machine then moves to the next iteration.

@@

The operation of such a machine is significantly more complex than the earlier one - a full explanation of its operation can be found in appendix \ref{appendix:turingexampleexp}. Key to the operation is the fact that the tape begins with the sequence \texttt{e e 0} at the start of the tape (this method shall be used in future), and that cells are seperated into value cells and marker cells. On value cells (odd-numbered cells) the symbols of the output sequence are printed. Only the symbols \texttt{0} and \texttt{1} will ever be printed on these cells. To the right of each value cell is a label cell (an even-numbered cell). These cells will either be left empty or will contain a 'marker' symbol, which labels the symbol on the value cell to the left. These symbols can be erased and overwritten, and are not part of the output sequence. Only one label cell is needed to contain multiple markers, or more descriptive markers for a single value cell, as this can be achieved with a sufficiently rich variety of marker symbols. This method is frequently used by Alan Turing in \cite{turing1936computablenumbers}.

It is clear from this example that the operation of the Turing Machine is very different from that of an ordinary computer. However, the core components are still mostly present - conditionals in the form of the symbol tests in the second column, variables in the form of labels, and continuity and looping when a state triggers itself - a form of recursion. The end result is the same - a solution is produced in the form of the computable sequence written on the value squares, and the machine can continue this sequence without halting indefinitely. Unfortunately, the operation of this machine is somewhat complex and obfuscated. To simplify this, we will introduce a type of function for the Turing machine.

@@

\textit{M}-functions, described by Turing in \cite{turing1936computablenumbers}, are effectively an extension of states. They are denoted by a lowercase letter and a set of arguments contained in brackets - arguments in gothic script refer to states (or other \textit{m}-functions), and arguments in italics refer to symbols. When an \textit{m}-function is triggered, the italic and gothic symbols in the description are substituted by the supplied arguments (similar to application in $\lambda$-Calculus). An example of an \textit{m}-function would be the function $\mathbb{f}$ below: \cite{turing1936computablenumbers}

@@

The function $\mathbb{f}$ takes three arguments - the states \textfrak{C} and \textfrak{E} and the symbol \textit{a}. It is not difficult to see that if $\rightarrow \mathbb{f}$($\mathbb{A}$, $\mathbb{B}$, x) the cursor would trigger $\mathbb{A}$ at the first occurrence of \texttt{x}. If no symbols \texttt{x} were found, \textfrak{E} would be triggered on the marker of the first empty value square.

@@

\medskip\noindent\begin{tabu} to \textwidth{XXXXX}
    \multicolumn{2}{c}{\textit{Configuration}} & \multicolumn{2}{c}{\textit{Behavior}} \\
    \textit{state} & \textit{symbol} & \textit{transition\qquad function} & \textit{final state} \\
    \hhline{====}
    $\mathbb{e}$(\textit{a}, \textfrak{C}, \textfrak{E}) & - & - & $\mathbb{f}$(\textit{a}, $\mathbb{e}_1$(\textfrak{C}), \textfrak{E}) & \multirow{2}{80pt}{Erases first occurrence of symbol \textit{a}} \\
    \hhline{----}
    $\mathbb{e}_1$(\textfrak{C}, \textfrak{E}) & - & E & \textfrak{C}\\
    \hhline{====}
    $\mathbb{l}$(\textfrak{C}, \textfrak{E}) & - & L & \textfrak{C} & Moves cursor left\\
    \hhline{----}
    $\mathbb{r}$(\textfrak{C}, \textfrak{E}) & - & R & \textfrak{C} & Moves cursor right\\
    \hhline{====}
    $\mathbb{fl}$(\textit{a}, \textfrak{C}, \textfrak{E}) & - & - & $\mathbb{f}$(\textit{a}, $\mathbb{l}$(\textfrak{C}), \textfrak{E}) & Moves to left of \textit{a} \\
    \hhline{----}
    $\mathbb{fr}$(\textit{a}, \textfrak{C}, \textfrak{E}) & - & - & $\mathbb{f}$(\textit{a}, $\mathbb{r}$(\textfrak{C}), \textfrak{E}) & Moves to right of \textit{a} \\
    \hhline{====}
    $\mathbb{p0}$(\textfrak{C}, \textfrak{E}) & - & P\texttt{0} & \textfrak{C} & Prints \texttt{0} \\
    \hhline{----}
    $\mathbb{p1}$(\textfrak{C}, \textfrak{E}) & - & P\texttt{1} & \textfrak{C} & Prints \texttt{1} \\
    \hhline{----}
    $\mathbb{p}$(\textit{s},\textfrak{C}, \textfrak{E}) & - & P\textit{s} & \textfrak{C} & Prints \textit{s} \\
    \hhline{====}
    $\mathbb{n}$(\textfrak{C}, \textfrak{E}) & - & - & $\mathbb{f}$(\texttt{e}, $\mathbb{n}_1$(\textfrak{C}), \textfrak{E}) & \multirow{3}{80pt}{Moves to first empty (final) \textit{p}-square} \\
    \hhline{----}
    \multirow{2}{*}{$\mathbb{n}_1$(\textfrak{C})} & Any  & R, R & $\mathbb{n}_1$(\textfrak{C}) \\
                                                  & None & - & \textfrak{C} \\
    \hhline{====}
\end{tabu}

@@

At this point, the similarity between Turing machines and modern computers becomes very clear. The machine is able to store information, from integers to strings of characters to images, as binary sequences of value square \texttt{0}s and \texttt{1}s - these shall be stored on the tape as memory. The first data structure we shall define is the simple boolean - only 1 bit (1 value square) long. In order to identify these sequences on the tape as booleans, we shall label the first character of the boolean with an identifier (we shall call these pointers) unique to that boolean variable - for example the symbol \texttt{\textbf{B}myvar} (bear in mind this is one symbol, not six) would mark a boolean named \textit{myvar}. Let us define some \textit{m}-functions to define and assign values to variables of the boolean type:

@@

\subsection{Scripted Configurations}
As we have moved into more and more complex Turing machine functions, it can be noted that the middle two columns of the description are becoming gradually more empty - this is due to the fact that their operation can be entirely represented within \textit{m}-functions. We have started doing so with the functions $\mathbb{l}$, $\mathbb{r}$, $\mathbb{p0}$, and $\mathbb{p1}$ - furthermore, it is not difficult to create conditional functions to replicate the operation of the second column.

@@

The function $\mathbb{if}$ takes a symbol argument, which is compared to the symbol under the cursor - if they are the same, the machine $\rightarrow$ \textfrak{I}; if not the machine $\rightarrow$ \textfrak{O}.
Since we have now entirely defined the operation of the machine in terms of \textit{m}-functions, we no longer have any need for the central two table columns. In continuing to define \textit{m}-functions, we could simply choose to omit these columns, or even omit the table altogether and use a clearer notation - in this case, for a definition of the if function but based on pointers:

@@

\subsection{Universality}

Turing machines are truly remarkable entities. While $\lambda$-Calculus was able to mimic the operation of a modern computer (behaving like a purely functional programming language such as Haskell), its axioms were completely abstract - the conceptual operations of abstraction and application, while being probably more useful than Turing machines in terms of yielding practical results, were unlike anything which could actually be created in the real world. However, Turing machines are entirely grounded in physical reality, and it is totally conceivable how one could create such a machine, while still being able to produce results very similar to that of an imperative programming language such as C - data storage, typed and referenceable variables, and boolean logic, all performed as a sequence of commands. From these building blocks, integers and more complex concepts can be produced. This is the truly remarkable thing about Turing machines - they illustrate clearly how mechanisms can in fact solve computable problems.

However, up to this point, computational ability has been measured by the potential of a machine to perform a variety of tasks that appear to demonstrate computation. This is not a sufficiently rigorous definition - a machine which seems to behave like a computer is not necessarily able to evaluate any computable function. For practicality's sake, it seems natural for Turing Machines to be chosen to be the pinnacle of computability: any machine that can be programmed in such a way that it emulates a Turing machine shall be called a  `Turing-complete' computer, and any problem a Turing machine can solve is a computable function.

It is unseemly that the abstract notion of computability relies necessarily on the specific design of the Turing machine: the idea of computability is a robust, almost Platonic reality, that Turing machines, $\lambda$-Calculus and all other effectively calculable devices, physical or theoretical, participate within equally. In fact, the idea of Turing-completeness presented earlier achieves this aim. This is the concept of Universality.

The main idea behind Universality is the concept of the Universal Machine, a machine which is able to entirely emulate another given system. As shown in \cite{turing1936computablenumbers}, a Turing machine can be configured in such a way that it emulates a Turing machine; this machine reads a description table from the start of the tape, and then produces the output that the machine described in the pre-written table would produce. Furthermore, $\lambda$-Calculus is also Turing-complete - this is demonstrated entirely in appendix \ref{appendix:lambdauniversaldemo}. This is done by producing a function that takes the current configuration of a Turing machine (in terms of the tape, state and description table) as input, and transforms it into the next configuration in sequence that a Turing machine would produce. This function would then be run repeatedly on the configuration using the \textbf{Y} combinator, hence emulating the action of Turing machines. Hence the $\lambda$-Calculus can produce a Universal Machine for Turing machines.

Most remarkably, Turing machines are in fact able to also emulate $\lambda$-Caluculus. This shall not be demonstrated directly, but it seems intuitive that a Turing machine could carry out processes of abstraction and application since they are at heart simple algorithmic processes; a Turing macihne would evaluate $\lambda$-terms in a similar way to how the human mind would. This leads us onto the remarkable fact that since Turing machines and $\lambda$-Calculus can produce Universal Machines for each other, each has the potential to perform any task the other can perform, and hence they are equal in power. Furthermore, many other theoretical methods of computation such as Kurt G\"odel's General Recursive Functions also share this equivalence \cite{copeland2002ctt}.

It therefore seems that the informal notion of computability can be formally defined with the use of one of these equivalent computational systems - this is a truly remarkable result, which produces a very strong basis for computability. With this test for computability in place, the question of whether World War 2 produced computing machines becomes apparently trivial.
